---
title: "Estimating effective reproductive number using clinical report data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rt_clinical_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressMessages({
  library(magrittr)
  library(tidyr)
  library(dplyr)
  library(ggplot2) ; theme_set(theme_bw())
  library(lubridate)
  library(stringr)
  library(patchwork)
  library(ern)
})
```

With the `ern` package, the effective reproduction number ($\mathcal{R}_t$) can be estimated using clinical reports, aggregated over time (_e.g._ weekly, biweekly). In this vignette, $\mathcal{R}_t$ will be estimated using a subset of weekly COVID-19 reports from the Government of Canada Health Infobase dataset. 

## Loading data and necessary parameters

In order to estimate $\mathcal{R}_t$ from aggregated clinical testing data in `ern`, the following items are required:

- ...

Sample weekly clinical reports are included in `ern`:

```{r load-data}
cl.agg <- ern::cl.agg
```

